Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2023-03-31 src/tex/ms.tex
Chat

Since we know intrinsic parameters play an important role in the ansatz, we
would like to investigate how intrinsic parameters affect the recalibration
process. First, we plot the parameter space of $q$ vs. $\chi_{\mathrm{PN}}$ in
Fig. \ref{fig:ps_q148_qchi}. In the low mass ratio region, waveforms with both
positive and negative log differences are mixed up. On the other hand, along the
horizontal line $\chi_{\mathrm{PN}}=0$, waveform mismatches consistently improve
under recalibration.

Furthermore, we plot the parameter space of $\chi_1$ vs. $\chi_2$ in
Fig.~\ref{fig:ps_q148}. Waveforms along the diagonal axis, i.e.
$\chi_1\approx\chi_2$, show good mismatch improvements. We expect to see this
feature since the original coefficients were fitted using NR waveforms with
equal or similar spin, hence the model prefers waveforms with similar spin. One
interesting feature is how the second and fourth quadrants respond to
optimization. In the second quadrant ($\chi_1<0$ and $\chi_2>0$), waveforms
generally improve with along optimization. However, mismatches in the fourth
quadrant ($\chi_1>0$ and $\chi_2<0$) do not improve after optimization. Most
waveforms even turned worse after optimization. These waveforms correspond to
the waveforms in the high mismatch tail in Fig.~\ref{fig:q148}.

% Waveform mismatches in the fourth quadrant, i.e. $\chi_1>0$ and $\chi_2<0$,
% performed worse after optimized. IMRPhenomD is constructed using single spin
% parameter, where there are degeneracy between $\chi_1$ and $\chi_2$. In
% regions with opposite extreme spin components, degeneracy significantly alters
% intrinsic parameters passed into waveform generation, hence the fourth
% quadrant shows such feature.

@@

As we see in Fig.~\ref{fig:ps_q148} that the recalibration procedure is
significantly different in different regions in the parameter space, we split
the parameter space into 4 quadrants and perform separate fitting with training
waveforms listed in Table \ref{tab:quadrants}. Note that in the second and
fourth quadrants, there are not enough waveforms with $q>4$, hence the result
are only valid up to $q\leq4$. From Fig.~\ref{fig:ps_q148_quadrant}, we see that
all waveforms, except those in the fourth quadrant, show improvement in
mismatch. Many features seen in Fig.~\ref{fig:ps_q148} can be found here again.
First, we focus on waveforms with the same spin direction, i.e.
$\chi_1\chi_2>0$. Waveforms lying in the neighborhood of the diagonal axis have
noticeable improvements in mismatch due to most training waveforms being
equal-spin waveforms. For the second quadrant, waveforms improved significantly
with only a few defects due to some testing waveforms having $q>4$. In the
fourth quadrant, most optimized waveforms have a higher mismatch than the
original waveforms, as indicated by the positive log difference of mismatches.
As performing optimization in a smaller subspace does not show any improvement
compared to the original result, the ansatz mostly does not fit waveforms in the
fourth quadrant.

@@

\begin{figure}[t]
	\script{ps_q148_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/ps_q148_quadrants.pdf}
	\caption{Parameter space of testing waveforms. Each quadrant is fitted independently. Colorbar represents log difference of mismatches before and after optimization.}
	\label{fig:ps_q148_quadrant}
\end{figure}

@@

We see that in Fig.~\ref{fig:q148_q1248_compare}, performing optimization with
more training waveforms only has a small increase in the accuracy of the
waveform model. We believe that by increasing the number of waveforms, the
accuracy will not have a significant change as the waveform coefficients are
already over-determined. Using more calibration NR waveforms will not further
improve the model significantly. This suggests the form of the parameterized
ansatz is not suitable for certain regions in the parameter space, thus
mismatches of only a few waveforms decreased while other waveforms remain at the
high mismatch tail with little changes. This implies that the model is
ultimately restricted by the flexibility of the ansatz.

One of the major problems causing inaccuracy in the ansatz is the reduced spin
approximation. In IMRPhenomD, it is modeled using a single spin parameter,
namely $\chi_{\mathrm{PN}}$ as outlined in Sec. \ref{sec:method}. Parameterizing
a BBH merger with one spin parameter introduces degeneracy within the parameter
space. Events with distinct black hole spins could result in equal
$\chi_{\mathrm{PN}}$, thus generating the same waveform. Especially with high
unequal spin events, $\chi_{\mathrm{PN}}$ would identify them as the same as
events with small spin, thus giving an inaccurate result. Generally, the
approximation gives straight lines of degeneracy in the parameter space, with
its slope (always negative) dependent on the mass ratio. From
Fig.~\ref{fig:ps_q148}, we see that along a degeneracy line, the ansatz behaves
better in the top left region than the bottom right. To try to accommodate this
issue, we split the parameter space into 4 quadrants as described in Sec.
\ref{sec:result}. However, even with separate optimizations, we see in
Fig.~\ref{fig:ps_q148_quadrant} that the fourth quadrant still shows similar
mismatches as before while the second quadrant further improved. This suggests
the ansatz is region-specific, with a higher preference for BBH events with
$\chi_1<0$ and $\chi_2>0$.

Separating regions into 4 quadrants is done purely out of simplicity. To give a
more comprehensive analysis, one should be systematic about region selection.
One can use level set estimation algorithms to obtain systematic regions of
interest. This general algorithm reveals further degeneracies or issues within
the ansatz. Then, recalibrating such individual regions might give better
results. Alternatively, one can select regions according to the lines of
degeneracy. However, with limited NR waveforms, such a selection scheme is not
viable for us. In the future, with more NR waveforms spanning the entire
parameter space, one can perform optimization with fewer restrictions.

While our work focused mainly on the IMRPhenomD model, this simple yet general
method can be utilized in other differentiable GW models. For instance, within
the same family, IMRPhenomX \citep{pratten2020setting} or IMRPhenomP
\citep{hannam2014simple} models. By jointly fitting a new set of coefficients,
it is expected that both models can be improved since they are constructed in a
similar way as the IMRPhenomD model. For example, they also use PN approximant
as part of the ansatz in the inspiral segment. One interesting result might
arise while recalibrating IMRPhenomX model \citep{pratten2020setting}. Since it
is parameterized by an additional anti-symmetric spin parameter, it is expected
to not show the same degeneracy as described above. Further analysis might give
insights into the systematics of Phenom models. Moreover, this method can be
applied to other GW model families, such as NR surrogate models
\citep{varma2019surrogate} or EOB models \citep{taracchini2014effective}. NR
waveform calibration procedures could be made easier and are likely to improve
current models.

>>>>

Since we know intrinsic parameters play an important role in the ansatz, we
would like to investigate how intrinsic parameters affect the recalibration
process. First, we plot the parameter space of $q$ vs. $\chi_{\mathrm{PN}}$ in
Fig. \ref{fig:ps_q148_qchi}. We can see near non-spinning waveforms demonstrate
more consistent improvement, whereas spinning waveforms can either improve or
worsen in terms of mismatch. \kw{What does this means?}

Furthermore, we plot the parameter space of $\chi_1$ vs. $\chi_2$ in
Fig.~\ref{fig:ps_q148}. Waveforms along the diagonal axis, i.e.
$\chi_1\approx\chi_2$, show good mismatch improvements. We expect to see this
feature since the original coefficients were fitted using NR waveforms with
equal or similar spin, hence the model prefers waveforms with similar spin. One
interesting feature is how the second and fourth quadrants respond to
optimization. \kw{How did you count quadrant?} In the second quadrant ($\chi_1<0$ and $\chi_2>0$), waveforms
generally improve with along optimization. However, mismatches in the fourth
quadrant ($\chi_1>0$ and $\chi_2<0$) do not improve after optimization. Most
waveforms even turned worse after optimization. These waveforms correspond to
the waveforms in the high mismatch tail in Fig.~\ref{fig:q148}. \kw{Give justification.}

@@

We divided the parameter space into four quadrants to analyze the effect of the
recalibration procedure on each region separately (Fig.\ref{fig:ps_q148}). The
training waveforms used for fitting are listed in Table \ref{tab:quadrants}. The
second and fourth quadrants have limited data for $q>4$, hence the results are
only valid up to $q\leq4$. From Fig.\ref{fig:ps_q148_quadrant}, we observe
improvements in waveform mismatch for all waveforms, except those in the fourth
quadrant. For equal-spin waveforms lying in the proximity of the diagonal axis,
the mismatch improves significantly due to most training waveforms being
equal-spin. In the second quadrant, the improvement is noticeable, except for a
few defects caused by some testing waveforms having $q>4$. The optimized
waveforms in the fourth quadrant have a higher mismatch than the original ones,
indicating poor fitting. Further optimization in a smaller subspace does not
improve the results.

@@

\begin{figure}[t]
	\script{ps_q148_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/ps_q148_quadrants.pdf}
	\caption{\kw{The color scheme makes the plot a bit hard to read.}Parameter
	space of testing waveforms. Each quadrant is fitted independently. Colorbar
	represents log difference of mismatches before and after optimization.}
	\label{fig:ps_q148_quadrant}
\end{figure}

@@

The results presented in Fig.~\ref{fig:q148_q1248_compare} demonstrate that
increasing the number of training waveforms used in waveform optimization yields
only a marginal increase in accuracy. Our analysis suggests that this marginal
improvement is a consequence of over-determination of the waveform coefficients.
Consequently, increasing the number of calibration NR waveforms is unlikely to
result in any significant improvement of the model's accuracy. These
observations suggest that the parameterized ansatz employed in our study may not
be suitable for certain regions in the parameter space, leading to mismatches
for some waveforms while other waveforms remain at the high mismatch tail with
negligible changes. This highlights the constraints of the model's flexibility
that ultimately limit its performance.

The reduced spin approximation is a major contributor to the inaccuracies
observed in the ansatz. In IMRPhenomD, this approximation employs a single spin
parameter, $\chi_{\mathrm{PN}}$, as described in Sec. \ref{sec:method}. The
parameterization of BBH mergers using a single spin parameter results in a
degeneracy within the parameter space. Specifically, black hole events with
different spins may generate the same waveform due to identical values of
$\chi_{\mathrm{PN}}$, leading to erroneous results, particularly for highly
unequal spin events. This degeneracy produces straight lines in the parameter
space with negative slopes that depend on the mass ratio, as shown in
Fig.\ref{fig:ps_q148}. Notably, the ansatz performs better in the top left
region than the bottom right along a degeneracy line. In an attempt to address
this issue, we partitioned the parameter space into four quadrants, as described
in Sec. \ref{sec:result}. However, even with separate optimizations for each
quadrant, Fig.\ref{fig:ps_q148_quadrant} indicates that the fourth quadrant
exhibits similar mismatches as before, while the second quadrant's performance
has improved. This observation suggests that the ansatz is specific to certain
regions of the parameter space, with a preference for BBH events featuring
$\chi_1<0$ and $\chi_2>0$.

The division of the parameter space into four quadrants was a simple approach
taken for practical reasons. A more systematic approach would involve the use of
level set estimation algorithms to identify regions of interest within the
parameter space. Such an algorithm can reveal additional degeneracies or issues
that may exist within the ansatz. One possible strategy is to recalibrate
individual regions of interest to achieve better results. An alternative
approach is to select regions based on degeneracy lines. However, due to the
limited number of NR waveforms available, we were unable to implement this
approach. With more NR waveforms available in the future that cover the entire
parameter space, we can perform optimization with fewer restrictions and select
regions more systematically.

Although our study primarily focused on the IMRPhenomD model, this simple yet
versatile approach can be applied to other differentiable GW models, such as the
IMRPhenomP \citep{hannam2014simple} or IMRPhenomX \citep{pratten2020setting} 
models within the same family. By jointly optimizing a new set of coefficients,
it is expected that both models can be enhanced since they share similar
construction principles to the IMRPhenomD model. For instance, they also use PN
approximants as part of the ansatz in the inspiral segment. Interestingly, while
recalibrating the IMRPhenomX model \citep{pratten2020setting}, which is
parameterized by an additional anti-symmetric spin parameter, it is expected not
to exhibit the degeneracy previously described. A more detailed investigation
may provide valuable insights into the systematics of the Phenom models.
Furthermore, this approach is applicable to other GW model families, such as NR
surrogate models \citep{varma2019surrogate} or EOB models
\citep{taracchini2014effective}. Such an approach could simplify NR waveform
calibration procedures and lead to the improvement of existing models.
